Retirement Consumption: Evidence from UK Pension Reform

### 2024-11-21 Texfiles/5348H.tex
Mostly stlyer changes to R scripts

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[authoryear]{natbib}
\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{siunitx}
\newcolumntype{d}{S[ input-open-uncertainty=, input-close-uncertainty=,
      parse-numbers = false, table-align-text-pre=false,
      table-align-text-post=false ]}
\usepackage{pdflscape}
\usepackage{hyperref}

@@

\begin{document}
\begin{titlepage}
    \begin{center}

@@

        Word count: 8336\footnote{This is an estimate that my \LaTeX editor gave me.} + 1650 (figure and table word equivalent)
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \citep{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts -- this phenomenon has been termed the ``annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature.

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot -- common advice is to take 4\% a year -- whilst the
rest remains invested. Subsequently, the number of annuities sold in the UK
dropped precipitously.

@@

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. A bequest motive means that individuals want to
leave an estate for their heirs to inherit -- which they cannot do if they
annuitise their wealth. On the other hand, individuals may believe that they
will not live as long as annuity providers think they will, and therefore
annuities that are on the market do not appear to be good value. Depending on
the reason for the lack of annuitisation in the UK, the consumption response of
retirees to the pension reform will differ. If individuals do not annuitise
because of pessimistic life expectancy, their consumption should increase after
the reform. If, on the other hand, individuals do not annuitise because of a
bequest motive, consumption should not change as much as a result of the reform.

@@

The importance of retirement policy to the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS), and for every 1,000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
\citep{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age \citep{cribb_karjalainen_ifs_2023}. Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals may require expensive end of life care. On the
other hand, if retirees over-save and do not spend, perhaps in order to bequest
wealth, the economy may be dynamically inefficient -- a state in which the
capital stock is too high. In this case consumption, per capita could be
increased if savings were decreased.

@@

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement, so
that they are insured against the risk of long life. More recently,
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then, there has been much
discussion of possible reasons for why people choose not to annuitise.
\cite{finkelstein_porteba_2002, finkelstein_porteba_2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell_et_al_1999} find
evidence of the same in the US. This means longer-lived individuals select into
buying annuities, driving up annuity prices since providers need to pay
annuitants for longer than they originally predicted. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell_et_al_1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (meaning the present discounted value of annuities
equals the price that an individual pays for one).

However, \cite{friedman_warshawsky_qje_1990} show that annuitisation decisions
can be fully explained by social security and `actuarially unfair' annuities, at
least for decisions in early retirement. They solve an augmented lifecycle model
and calculate annuity demand for a range of money's worths. For plausible values,
they find that individuals would optimally not annuitise much wealth. Similarly,
to \cite{finkelstein_porteba_2004}, \cite{friedman_warshawsky_chicago_1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But, they find this cannot fully explain
the annuitisation puzzle, and only when bequest motives are added to the model
can annuitisation rates be rationalised.

More recently, \cite{lockwood_red_2012} builds on this by showing that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks match other important aspects of the empirical
distribution of bequests, such as how much individuals bequeath and the
financial characteristics of these individuals. \cite{lockwood_aer_2018} also
finds that a model that treats bequests as luxury goods fits the lack of demand
for long-term care insurance well. \cite{odea_sturrock_rest_2023} find that
subjective life expectancies can explain a large part of the annuitisation
puzzle, especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Precautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than
an annuity. Moreover, \cite{vidalmelia_lejarragagarcia_munich_2004} suggest
that couples might share the risk of long life by pooling their resources, and
therefore couples may be less likely to annuitise their wealth. Another
explanation is that individuals already have enough guaranteed income from DB
pension schemes and state pensions that they do not need to annuitise more.

@@

The main dataset I use is the English Longitudinal Study of Ageing (ELSA)
\citep{main_elsa_citation}. ELSA picks individuals over the age of 50 to survey
every two years until death. If individuals leave the survey, ELSA replaces them
so that it is representative of the over 50 population in the UK. Individuals
are asked a range of questions relating to their income and wealth as well as
expectations about the future. One benefit of using ELSA is that it includes
data on pension types for individuals who are working. Therefore, I can
differentiate between individuals who have a DC pension and those who have a DB
pension. There have been 9 waves of ELSA data collection -- the first was in
2002/03, and collections happened every two years thereafter.

@@

I also use `life tables' from the UK's ONS from 2014. These provide me with the
risk of death for each age group. The tables are produced until age 120 and I
adjust them to make death certain at age 110 as is common in the literature
\citep{odea_sturrock_rest_2023}. I transform these so that I have risk of death
conditional on being a given age since this is what I use in the life cycle
simulations. I also use these objective probabilities to calculate annuity
prices for individuals.

@@

To calculate subjective life expectancies from ELSA data, I follow
\cite{odea_sturrock_rest_2023}. Individuals were asked “What are the chances
that you will live to be age X or more?” where X changed depending on the age of
the interviewee. If individuals were under 65 then X was 75; if individuals were
66 and older they were asked the age that was 11 to 15 years older than them and
a multiple of 5. From wave 3, respondents were also asked “What are the chances
that you will live to be age 85 or more?” if they were under 70. As most recent
retirees are under 70 we therefore have two data points. I first drop from the
data individuals who think it is more likely that they reach a higher age than a
younger age since this shows a misunderstanding of the question. I then add as a
third data point their objective chance of reaching 110 according to the ONS
life tables. I fit these three points to a Weibull distribution, which is
commonly used by demographers to estimate how populations will age, using
non-linear least squares. Then I create subjective survival tables using
parameters from the Weibull distribution.

@@

In the June 2010 budget, the coalition government announced the first of two key
reforms it would make to the pension system between 2010 and 2015. The 2010
policy reform created a minimum income requirement above which individuals would
not need to annuitise \citep{finance_act_hmt_2011}, meaning high income
retirees would not need to annuitise their wealth. The income requirement was
set at £20,000. This meant, for example, that an individual who was a member of
a DB scheme paying them £20,000 a year would not need to annuitise their DC
pension pot. However, the relatively high minimum income requirement meant that
most individuals still had to purchase an annuity.

In spring 2014, Osbourne, the chancellor at the time, announced the so called
`pensions freedom act', scrapping the minimum income requirement and eliminating
the compulsory annuities market \citep{pen_freedoms_hmt_2014}. One government
minister infamously brought the message home by saying pension pots `can be used
to buy Lamborghinis' if retirees wanted to \citep{guardian_lambos}. The
government also announced plans to make switches between DB and DC pension
schemes possible although this was only realised in the next parliament.

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Figure \ref{fig:annovertime} shows
purchases of annuities over time and demonstrates the sharp decrease in purchases
that ocurred from 2014 to 2015. There was an increase in the number of pensions
that were being accessed using an income drawdown product, but these only partly
account for the drop in annuitisation as is shown by the graph.

@@

    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_pot_sizes.pdf}
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \label{fig:ann2122}

    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution of methods used to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most DC pension pots were accessed through
annuitisation.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if this was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK --
changing the money's worth of an annuity. I run a robustness test in the
empirical section to check whether this reform biases my results.

@@

Table \ref{tab:sum_stats} shows various summary statistics for each variable of
interest from ELSA and the ONS.

@@

The treatment group has higher financial wealth with a median of £67,000 at the
time of interview as opposed to £54,300 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
more money in them. Both groups are equally likely to have a DB pension with
roughly 45\% of individuals across both samples having a DB pension. In general,
DB pensions are more prevalent than DC pensions in the data -- this reflects the
trend that is observed in the general UK population. Home ownership is roughly
equal across groups although the second group has slightly higher housing
wealth.

@@

\begin{landscape}
    \linespread{1.25}
    \input{tables/r_summary_stats.tex}
    \normalsize
\end{landscape}

@@

In particular I run:
\begin{equation*}
    Y_{i} = \alpha + \beta PostRef_{i}  + \epsilon_{i}
\end{equation*}

@@

Table \ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0, which shows that individuals are not
delaying retirement, beyond their original expectation, in order to be part of
the treatment. DC pension value is higher, as is financial wealth but these are
both imprecisely estimated. Because of the differences in these variables I add
them to the regression model.

One threat to validity is manipulation of the running variable, retirement year.
As observed above in Tables \ref{tab:sum_stats} and \ref{tab:cov_balance}, it
does not appear as though individuals are delaying retirement. However, it is
possible that individuals do not buy an annuity at the time of retirement since
the law prior to 2014 only required that they access the pot by age 75. They may
decide to keep their DC pension untouched and live off other income and assets
before accessing it later. Since I do not have data for when individuals
purchase annuities, I cannot track whether annuity purchases were delayed for
people in the treatment group. However, there is relatively stable annuity
demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so that they would not have to buy an annuity at all, we would observe declines
in purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a spike in annuity purchases prior to 2014.

@@

I use a fuzzy regression discontinuity design for which I compare the
consumption of individuals who retired after the policy reform to that of people
who retired just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing
the change as a ``bombshell" \citep{money_marketing_announcement}.

@@

\begin{equation*}
    Cons_{it} =  \gamma X_{it} + \beta PostReform_{it} + \epsilon
\end{equation*}

Where $Cons_{it}$ is one of five different consumption variables and $X_{it}$ is
a set of controls.

Table \ref{tab:DcOnlyRes} shows the results of this specification on the various
consumption indicators. Column 1 has total monthly consumption on the left hand
side of the regression equation. We can see that being in the treatment group is
associated with £36 more a week in spending overall. This is statistically
significant using robust standard errors clustered at the treatment level. Being
further into retirement at the time of interview is associated with higher total
consumption -- this is surprising since lifecycle models predict consumption
decreases over retirement and this has also been documented empirically
\citep{hurd_rohwedder_nber_2003}.

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This wealth effect may be driving the
results we observe in Table \ref{tab:DcOnlyRes}, since food consumption outside the
home increases as does expenditure on clothing -- seen in column 5.

One concern is that the group who retired after 2014 are just generally
different to those who retired before and perhaps consume more outside the home
anyway. The financial characteristic data presented in Table \ref{tab:sum_stats}
would support this hypothesis as it shows all retirees are richer post-reform.
To test for these cohort effects, I run a difference-in-difference (DD)
regression to complement the regression discontinuity just shown. I add
individuals who have a DB pension and those with no private pension into the
model and interact the policy change with dummies for having different types of
pension. Specifically, I run:

\begin{equation*}
    Cons_{it} =  \gamma X_{it} + \beta_{1} PostReform_{it} + \beta_{2} PostReform*DC_{it} + \beta_{3} PostReform*DB_{it}  + \epsilon
\end{equation*}

@@

Table \ref{tab:ElsaAllData} shows these results. The interactions at the bottom
of the table show us what happened to individuals with a given pension type
relative to the reference cohort, which has no pension. Those with only a DC
pension had an increase in total consumption which was higher than those with a
DB pension and higher than those with no pension. Consumption on food decreased
slightly for the DC group relative to no pension, but the decrease was less than
that for the DB group. Expenditure on clothing increased for the DC group
relative to the no pension and DB pension groups. The increase in sample size
also brings decreases in the standard errors of these variables -- the
interactions are now all significant a the 5\% level as is the main coefficient
on retiring post-reform.

@@

We can also test whether treatment intensity is correlated with larger increases
in spending. Those who have more money in a DC pension would be impacted by the
policy to a greater extent than those with smaller pensions, since they would be
forced to annuitise a larger amount. So, I interact the policy with DC wealth.
As in Table \ref{tab:DcOnlyRes}, this regression only uses those individuals who
have a DC pension.

@@

Table \ref{tab:DcOnlyInteract} shows the results. The interaction with DC
pension size is positive, meaning that a larger DC pension pot is associated with a
greater increase in total monthly consumption. For the all other categories,
more pension wealth is associated with lower consumption, though for food out
the house and for clothing the main coefficients are still positive and
significant.

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[style=authoryear]{biblatex}
\addbibresource{references.bib} %Import the bibliography file

\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{siunitx}
\newcolumntype{d}{S[ input-open-uncertainty=, input-close-uncertainty=,
      parse-numbers = false, table-align-text-pre=false,
      table-align-text-post=false ]}
\usepackage{pdflscape}
\usepackage{hyperref}

@@

\begin{document}

\begin{titlepage}
  \begin{center}

@@

    Word count: 8336\footnote{This is an estimate that my \LaTeX editor gave me.} + 1650 (figure and table word equivalent)
    \vfill
    I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University's and Faculty's definition of Plagiarism.
  \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
  I identify the consumption responses of retirees to the UK's `pensions
  freedom act', a landmark law change that meant defined contribution pension
  holders were no longer forced to annuitise their funds and could instead
  access their pensions in a variety of ways. I use a discontinuity in year of
  retirement and a difference-in-difference approach, harnessing those who
  were unaffected by the policy reform as a natural control group. I then
  solve three sets of lifecycle models, tracking assets from retirement until an
  agent reaches age 110. I simulate their consumption with and without
  annuities, mimicking the policy reform, and match their consumption
  responses to what was seen in the empirical models. I find that a lifecycle
  model with a bequest motive best fits the identified consumption response —
  a result that could inform policymakers in designing optimal pension
  policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death (\cite{yaari-65}). However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts --- this phenomenon has been termed the ``annuity puzzle''. The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature.

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010--2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot---common advice is to take 4\% a year---whilst the
rest remains invested. Subsequently, the number of annuities sold in the UK
dropped precipitously.

@@

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. A bequest motive means that individuals want to
leave an estate for their heirs to inherit --- which they cannot do if they
annuitise their wealth. On the other hand, individuals may believe that they
will not live as long as annuity providers think they will, and therefore
annuities that are on the market do not appear to be good value. Depending on
the reason for the lack of annuitisation in the UK, the consumption response of
retirees to the pension reform will differ. If individuals do not annuitise
because of pessimistic life expectancy, their consumption should increase after
the reform. If, on the other hand, individuals do not annuitise because of a
bequest motive, consumption should not change as much as a result of the reform.

@@

The importance of retirement policy to the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS), and for every 1,000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
(\cite{ons-population-predictions-2020}). The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age (\cite{cribb-karjalainen-ifs-2023}). Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals may require expensive end of life care. On the
other hand, if retirees over-save and do not spend, perhaps in order to bequest
wealth, the economy may be dynamically inefficient --- a state in which the
capital stock is too high. In this case consumption, per capita could be
increased if savings were decreased.

@@

\cite{yaari-65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement, so
that they are insured against the risk of long life. More recently,
\cite{davidoff-brown-diamond-aer-2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then, there has been much
discussion of possible reasons for why people choose not to annuitise.
\cite{finkelstein-porteba-2002, finkelstein-porteba-2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell-et-al-1999} find
evidence of the same in the US. This means longer-lived individuals select into
buying annuities, driving up annuity prices since providers need to pay
annuitants for longer than they originally predicted. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell-et-al-1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (meaning the present discounted value of annuities
equals the price that an individual pays for one).

However, \cite{friedman-warshawsky-qje-1990} show that annuitisation decisions
can be fully explained by social security and `actuarially unfair' annuities, at
least for decisions in early retirement. They solve an augmented lifecycle model
and calculate annuity demand for a range of money's worths. For plausible values,
they find that individuals would optimally not annuitise much wealth. Similarly,
to \cite{finkelstein-porteba-2004}, \cite{friedman-warshawsky-chicago-1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But, they find this cannot fully explain
the annuitisation puzzle, and only when bequest motives are added to the model
can annuitisation rates be rationalised.

More recently, \cite{lockwood-red-2012} builds on this by showing that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks match other important aspects of the empirical
distribution of bequests, such as how much individuals bequeath and the
financial characteristics of these individuals. \cite{lockwood-aer-2018} also
finds that a model that treats bequests as luxury goods fits the lack of demand
for long-term care insurance well. \cite{odea-sturrock-rest-2023} find that
subjective life expectancies can explain a large part of the annuitisation
puzzle, especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Precautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than
an annuity. Moreover, \cite{vidalmelia-lejarragagarcia-munich-2004} suggest
that couples might share the risk of long life by pooling their resources, and
therefore couples may be less likely to annuitise their wealth. Another
explanation is that individuals already have enough guaranteed income from DB
pension schemes and state pensions that they do not need to annuitise more.

@@

The main dataset I use is the English Longitudinal Study of Ageing (ELSA)
(\cite{main-elsa-citation}). ELSA picks individuals over the age of 50 to survey
every two years until death. If individuals leave the survey, ELSA replaces them
so that it is representative of the over 50 population in the UK. Individuals
are asked a range of questions relating to their income and wealth as well as
expectations about the future. One benefit of using ELSA is that it includes
data on pension types for individuals who are working. Therefore, I can
differentiate between individuals who have a DC pension and those who have a DB
pension. There have been 9 waves of ELSA data collection --- the first was in
2002/03, and collections happened every two years thereafter.

@@

I also use `life tables' from the UK's ONS from 2014. These provide me with the
risk of death for each age group. The tables are produced until age 120 and I
adjust them to make death certain at age 110 as is common in the literature
(\cite{odea-sturrock-rest-2023}). I transform these so that I have risk of death
conditional on being a given age since this is what I use in the life cycle
simulations. I also use these objective probabilities to calculate annuity
prices for individuals.

@@

To calculate subjective life expectancies from ELSA data, I follow
(\cite{odea-sturrock-rest-2023}). Individuals were asked “What are the chances
that you will live to be age X or more?” where X changed depending on the age of
the interviewee. If individuals were under 65 then X was 75; if individuals were
66 and older they were asked the age that was 11 to 15 years older than them and
a multiple of 5. From wave 3, respondents were also asked “What are the chances
that you will live to be age 85 or more?” if they were under 70. As most recent
retirees are under 70 we therefore have two data points. I first drop from the
data individuals who think it is more likely that they reach a higher age than a
younger age since this shows a misunderstanding of the question. I then add as a
third data point their objective chance of reaching 110 according to the ONS
life tables. I fit these three points to a Weibull distribution, which is
commonly used by demographers to estimate how populations will age, using
non-linear least squares. Then I create subjective survival tables using
parameters from the Weibull distribution.

@@

In the June 2010 budget, the coalition government announced the first of two key
reforms it would make to the pension system between 2010 and 2015. The 2010
policy reform created a minimum income requirement above which individuals would
not need to annuitise (\cite{finance-act-hmt-2011}), meaning high income
retirees would not need to annuitise their wealth. The income requirement was
set at £20,000. This meant, for example, that an individual who was a member of
a DB scheme paying them £20,000 a year would not need to annuitise their DC
pension pot. However, the relatively high minimum income requirement meant that
most individuals still had to purchase an annuity.

In spring 2014, Osbourne, the chancellor at the time, announced the so called
`pensions freedom act', scrapping the minimum income requirement and eliminating
the compulsory annuities market (\cite{pen-freedoms-hmt-2014}). One government
minister infamously brought the message home by saying pension pots `can be used
to buy Lamborghinis' if retirees wanted to (\cite{guardian-lambos}). The
government also announced plans to make switches between DB and DC pension
schemes possible although this was only realised in the next parliament.

The impact of the reforms on annuity demand has been documented by
\cite{cannon-et-al-nier-2016}. Figure \ref{fig:annovertime} shows
purchases of annuities over time and demonstrates the sharp decrease in purchases
that ocurred from 2014 to 2015. There was an increase in the number of pensions
that were being accessed using an income drawdown product, but these only partly
account for the drop in annuitisation as is shown by the graph.

@@

  \centering
  \includegraphics[width=0.9\columnwidth]{figures/annuity_pot_sizes.pdf}
  \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
  \label{fig:ann2122}
\end{figure}

Figure~\ref{fig:ann2122} shows the distribution of methods used to access pension pots
at retirement in 2021--22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case --- most DC pension pots were accessed through
annuitisation.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if this was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK ---
changing the money's worth of an annuity. I run a robustness test in the
empirical section to check whether this reform biases my results.

@@

Table~\ref{tab:sum_stats} shows various summary statistics for each variable of interest from ELSA and the ONS.

@@

The treatment group has higher financial wealth with a median of £67,000 at the
time of interview as opposed to £54,300 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
more money in them. Both groups are equally likely to have a DB pension with
roughly 45\% of individuals across both samples having a DB pension. In general,
DB pensions are more prevalent than DC pensions in the data --- this reflects the
trend that is observed in the general UK population. Home ownership is roughly
equal across groups although the second group has slightly higher housing
wealth.

@@

\begin{landscape}
  \linespread{1.25}
  \input{tables/r_summary_stats.tex}
  \normalsize
\end{landscape}
% \resizebox{\textwidth}{!}{  \input{tables/r_summary_stats.tex}}

@@

In particular I run:
\begin{equation*}
  Y_{i} = \alpha + \beta_{1} PostRef_{i} + \beta_{2} RetirementYear_{i}  + \beta_{2} RetirementYear_{i}PostRef_{i}  + \epsilon_{i}
\end{equation*}

@@

Table~\ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0, which shows that individuals are not
delaying retirement, beyond their original expectation, in order to be part of
the treatment. DC pension value is higher, as is financial wealth but these are
both imprecisely estimated. Because of the differences in these variables I add
them to the regression model.

One threat to validity is manipulation of the running variable, retirement year.
As observed above in Tables~\ref{tab:sum_stats} and~\ref{tab:cov_balance}, it
does not appear as though individuals are delaying retirement. However, it is
possible that individuals do not buy an annuity at the time of retirement since
the law prior to 2014 only required that they access the pot by age 75. They may
decide to keep their DC pension untouched and live off other income and assets
before accessing it later. Since I do not have data for when individuals
purchase annuities, I cannot track whether annuity purchases were delayed for
people in the treatment group. However, there is relatively stable annuity
demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so that they would not have to buy an annuity at all, we would observe declines
in purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a spike in annuity purchases prior to 2014.

@@

I use a fuzzy regression discontinuity design for which I compare the
consumption of individuals who retired after the policy reform to that of people
who retired just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing
the change as a ``bombshell'' (\cite{money-marketing-announcement}).

@@

\begin{equation*}
  \label{eq:main-rd}
  Y_{i} = \alpha + \beta_{1} PostRef_{i} + \beta_{2} RetirementYear_{i}  + \beta_{2} RetirementYear_{i} X PostRef_{i}  + \epsilon_{i}
\end{equation*}

Where $Y_{i}$ is one of five different consumption variables and $X_{i}$ is
a set of controls.

Table~\ref{tab:DcOnlyRes} shows the results of this specification on the various
consumption indicators. Column 1 has total monthly consumption on the left hand
side of the regression equation. We can see that being in the treatment group is
associated with £36 more a week in spending overall. This is statistically
significant using robust standard errors clustered at the treatment level. Being
further into retirement at the time of interview is associated with higher total
consumption---this is surprising since lifecycle models predict consumption
decreases over retirement and this has also been documented empirically
(\cite{hurd-rohwedder-nber-2003}).

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This wealth effect may be driving the
results we observe in Table~\ref{tab:DcOnlyRes}, since food consumption outside the
home increases as does expenditure on clothing---seen in column 5.

One concern is that the group who retired after 2014 are just generally
different to those who retired before and perhaps consume more outside the home
anyway. The financial characteristic data presented in Table~\ref{tab:sum_stats}
would support this hypothesis as it shows all retirees are richer post-reform.
To test for these cohort effects, I run a difference-in-difference (DD)
regression to complement the regression discontinuity just shown. I add
individuals who have a DB pension and those with no private pension into the
model and interact the policy change with dummies for having different types of
pension. Specifically, I run:

\begin{equation*}
  Cons_{i} =  \gamma X_{i} + \beta_{1} PostReform_{i} + \beta_{2} PostReform*DC_{i} + \beta_{3} PostReform*DB_{i}  + \epsilon
\end{equation*}

@@

Table~\ref{tab:ElsaAllData} shows these results. The interactions at the bottom
of the table show us what happened to individuals with a given pension type
relative to the reference cohort, which has no pension. Those with only a DC
pension had an increase in total consumption which was higher than those with a
DB pension and higher than those with no pension. Consumption on food decreased
slightly for the DC group relative to no pension, but the decrease was less than
that for the DB group. Expenditure on clothing increased for the DC group
relative to the no pension and DB pension groups. The increase in sample size
also brings decreases in the standard errors of these variables --- the
interactions are now all significant a the 5\% level as is the main coefficient
on retiring post-reform.

@@

We can also test whether treatment intensity is correlated with larger increases
in spending. Those who have more money in a DC pension would be impacted by the
policy to a greater extent than those with smaller pensions, since they would be
forced to annuitise a larger amount. So, I interact the policy with DC wealth.
As in Table~\ref{tab:DcOnlyRes}, this regression only uses those individuals who
have a DC pension.

@@

Table~\ref{tab:DcOnlyInteract} shows the results. The interaction with DC
pension size is positive, meaning that a larger DC pension pot is associated with a
greater increase in total monthly consumption. For the all other categories,
more pension wealth is associated with lower consumption, though for food out
the house and for clothing the main coefficients are still positive and
significant.

As mentioned above, the DC pension pot value data after
wave 5 has not been released yet, so for individuals who retire after 2011 I
have no DC wealth variable. For Table~\ref{tab:DcOnlyInteract}, pension wealth
from the last available year for an individual was given a real return of 3\%
and compounded until year of interview. Given that this is an imperfect measure,
I also run the regression with financial wealth interacted with the treatment
variable.
